Intrusion Detection with Genetic Algorithms and Fuzzy Logic

### 2013-10-30 paper.tex
Wrote more in the introduction.

In Section~\ref{sec:background} I will give some background on intrusion detection systems. I will explain the types of intrusion detection systems, the main types of networking attacks, and ways of detecting attacks. I will explain what \emph{rules} are in Section~\ref{sec:rules}, and then give information on fuzzy logic and genetic algorithms in Sections~\ref{sec:fuzzyLogic} and~\ref{sec:GA}. I will then explain two data sets that are used in intrusion detection. After that, I will explain the measures that are used in determining the accuracy of an intrusion detection system.

@@

\subsection{Fuzzy Logic}
\label{sec:fuzzyLogic}
Attacks on systems do not always have a fixed patter, so fuzzy logic is used to detect patterns that have a behavior that is between normal and unusual. A fuzzy logic rule has the following format: If <\emph{condition}> then <\emph{consequence}>, where \emph{condition} is a fuzzy variable, and \emph{consequence} is a fuzzy set. An example of this is: if the number of packets with the same destination address is high, then the pattern is unusual. To determine what is considered high, the values of the packets are divided into fuzzy sets. In Figure~\ref{fig:triangleFigure} there are three sets: low (region A), medium (region B), and high (region C). 
\begin{figure}
\centering
\psfig{file=triangleFigure.JPG}
\caption{Fuzzy shape with three sets: low (A), medium (B), and high (C)~\cite{DBLP:journals/corr/abs-1204-6416}}
\label{fig:triangleFigure}
\end{figure}
The x-axis are the values in the fuzzy set (number of packets), and the y-axis is the membership function, which determines the region. For example, if the number of packets with the same destination address is 15 then this will be considered as low for a degree of 0.4, but will be considered medium or high for other degrees. If the number of packets is determined to be high, then the connection will be aborted.~\cite{DBLP:journals/corr/abs-1204-6416}

\subsection{Genetic Algorithm}
\label{sec:GA}
Genetic algorithms are a search technique used to find solutions to problems \emph{Mutation}, \emph{selection}, and \emph{crossover} are used to evolve and improve rules.

Mutation is where random bits in a solution are changed. Selection is where solutions that have a better \emph{fitness} are chosen over the other solutions. The fitness function determines the quality of a particular solution. Crossover is where two solutions swap one of their characteristics with the other to form two new solutions.

@@

\begin{table}
\center
\caption{Number of records}
\begin{tabular}{lll}
  & Training & Testing \\ 
Normal & 97,280 & 60,593\\
Probe  & 4,107  & 4,166\\
DoS	   & 391,458 & 229,853\\
U2R    & 52      & 228\\
R2L    & 1,124   & 16,189\\
Total  & 494,021 & 311,029\\
\end{tabular}
\center
\label{tab:numberOfRecords}
\end{table}

\begin{table*}
\center
\caption{Results for Genetic Algorithm Experiment}
\begin{tabular}{cccccccc}
& & \multicolumn{5}{ c }{Predicted label} \\ 
& & Normal & Probe & DoS & U2R & R2L & \% Correct \\ \cline{3-8}
\multicolumn{1}{ c }{\multirow{2}{*}{Actual class} } &
\multicolumn{1}{ c| }{Normal} & 42138 & 1421 & 15835 & 486 & 713 & 69.5\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{Probe} & 398 & 2963 & 654 & 2 & 149 & 71.1   \\
\multicolumn{1}{ c  }{\multirow{2}{*}{} } &
\multicolumn{1}{ c| }{Dos} & 921 & 432 & 228489 & 1 & 10 & 99.4\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{U2R} & 146 & 21 & 8 & 43 & 10 & 18.9\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{R2L} & 11191 & 578 & 3398 & 141 & 881 & 5.4\\

@@

\begin{table*}
\center
\caption{Results for Genetic Algorithm Experiment}
\begin{tabular}{cccc}
%\cline{3-8}
& & \multicolumn{2}{ c }{Predicted label} \\ 
& & Normal & Intrusion\\ \cline{3-4}
\multicolumn{1}{ c }{\multirow{2}{*}{Actual class} } &
\multicolumn{1}{ c| }{Normal} & True Negative (42,138) & False Positive (18,455)\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{Probe} & False Negative (12,528) & True Positive (237,908)\\
\end{tabular}
\center
\label{tab:genAlgResults2}
\end{table*}

@@

\begin{figure}
\centering
\caption{Fuzzy encoding for a feature}
\begin{tabular}{|cccc|} \hline
010 & 011 & 100 & 101\\
a=2 & b=3 & c=4 & d=5\\
\hline\end{tabular}
\label{fig:fuzEncodingForFeature}
\end{figure}

\begin{figure*}
\centering
\caption{A rule with 12 blocks of features}
\begin{tabular}{|cccc|c|cccc|c|} \hline
010 & 011 & 100 & 101   & ...... & 010 & 011 & 101 & 111   & DoS\\
a=2 & b=3 & c=4 & d=5   & ...... & a=2 & b=3 & c=5 & d=7   &\\ 
    &     & Block 1&    &        &     & Block 12& &       & Type\\
\hline\end{tabular}
\label{fig:rule}
\end{figure*}

@@

\begin{table*}
\center
\caption{Results from Experiment 1}
\begin{tabular}{ccccccc}
 & Attack & Normal & Total Records & FP (\%) & FN (\%) & DR (\%)\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 16000 & 26500 & 1.13 & 4.10 & 97.92\\
\end{tabular}
\label{tab:fuzGenExp1}
\center
\end{table*}

In the second experiment, seven tests were run. For each test case there were 13 attack types plus normal activity that were in the training data set. Three attack types were used for the unknown testing data set. For example, test case 1 used the training data set that does not have Advance Port Scan, Ack Scan, and Xmas Tree, which are all probe attacks. These three attacks were then used for the testing data set. Table~\ref{tab:fuzGenExp2} shows some of the results from the fuzzy genetic algorithm and a decision tree algorithm, which is another common way of addressing these problems. For further information, see~\cite{decisionTree}. When compared with the decision tree algorithm, the fuzzy genetic algorithm has a higher detection rate in all cases except 3 and 5. It can be seen that in cases 2 and 4 the decision tree has low detection rates, while the fuzzy genetic algorithm has much higher detection rates.

\begin{table}
\caption{Unknown Attack Experiment}
\begin{tabular}{llll}
Test & Unknown & Decision & Fuzzy\\
Case & Attacks & Tree DR (\%)  & Genetic DR (\%)\\ \hline

@@

\begin{table}
\caption{KDD99 and RLD09 Results}
\begin{tabular}{cccccc}
Data set & Attack & Normal & FP (\%) & FN (\%) & DR (\%)\\ \hline
KDD99 & 160,117 & 39,337 & 0.13 & 1.55 & 98.72\\
RLD09 & 10,500 & 16,000 & 1.14 & 3.39 & 97.97\\
\end{tabular}
\label{tab:bothSetsResults}
\end{table}

@@

\begin{table}
\caption{Results for KDD99 with Certain Attacks}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & Back & DoS & 85.33 & 0.00 & 16.56\\
2 & Smurf & DoS & 0.76 & 0.10 & 99.73\\
3 & Portsweep & Probe & 6.40 & 0.00 & 93.66\\
4 & Satan & Probe & 0.74 & 3.75 & 99.22\\
\end{tabular}
\label{tab:kddAttacks}
\end{table}

@@

\begin{table}
\caption{Results for RLD99 with Certain Attacks}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & Smurf & DoS & 0.02 & 0 & 99.98\\
2 & UDP Flood & DoS & 11.06 & 0 & 89.59\\
3 & Ackscan & Probe & 0.03 & 0 & 99.97\\
4 & Synscan & Probe & 0.65 & 4.2 & 99.24\\
\end{tabular}
\label{tab:rldAttacks}
\end{table}

>>>>

In Section~\ref{sec:background} I will give some background on intrusion detection. I will explain the main types of networking attacks in Section~\ref{sec:typesAttacks}, and ways of detecting attacks in Section~\ref{sec:detectionMeth}. In Section~\ref{sec:dataSets} I will explain two data sets that are used in intrusion detection. I will explain what \emph{rules} are in Section~\ref{sec:rules}, and then give information on fuzzy logic and genetic algorithms in Sections~\ref{sec:fuzzyLogic} and~\ref{sec:GA}. After that, I will explain the measures that are used in determining the accuracy of an intrusion detection system in Section~\ref{sec:PosNeg}. In Section~\ref{sec:genAlgImp} I will explain the use of a genetic algorithm in an intrusion detection system, and in Section~\ref{sec:fuzGenAlgImp} I will explain the use of a fuzzy genetic algorithm. Section~\ref{sec:conclusion} has some conclusions about the use of genetic algorithms and fuzzy genetic algorithms in intrusion detection systems.

@@

\subsection{Fuzzy Logic}
\label{sec:fuzzyLogic}
Attacks on systems do not always have a fixed pattern, so fuzzy logic is used to detect patterns that have a behavior that is between normal and unusual. A fuzzy logic rule has the following format: If <\emph{condition}> then <\emph{consequence}>, where \emph{condition} is a fuzzy variable, and \emph{consequence} is a fuzzy set. An example of this is: if the number of packets with the same destination address is high, then the pattern is unusual. To determine what is considered high, the values of the packets are divided into fuzzy sets. In Figure~\ref{fig:triangleFigure} there are three sets: low (region A), medium (region B), and high (region C). 
\begin{figure}
\centering
\psfig{file=triangleFigure.JPG}
\caption{Fuzzy shape with three sets: low (A), medium (B), and high (C)~\cite{DBLP:journals/corr/abs-1204-6416}}
\label{fig:triangleFigure}
\end{figure}
The x-axis are the values in the fuzzy set (number of packets), and the y-axis is the membership function, which determines the region. For example, if the number of packets with the same destination address is 15 then this will be considered as low for a degree of 0.4, but will be considered medium or high for other degrees. If the number of packets is determined to be high, then the connection will be aborted.~\cite{DBLP:journals/corr/abs-1204-6416}

\vspace{0.30cm}

\subsection{Genetic Algorithm}
\label{sec:GA}
Genetic algorithms are a search technique used to find solutions to problems. \emph{Mutation}, \emph{selection}, and \emph{crossover} are used to evolve and improve rules.

Mutation is where random bits in an \emph{individual}, or possible solution, are changed. Selection is where individuals that have a better \emph{fitness} are chosen over the other individuals. The fitness function determines the quality of a particular individual. Crossover is where two individuals swap one of their characteristics with the other to form two new individuals.

@@

\begin{table}
\center
\caption{Number of records}
\vspace{0.20cm}
\begin{tabular}{lll}
  & Training & Testing \\ 
Normal & 97,280 & 60,593\\
Probe  & 4,107  & 4,166\\
DoS	   & 391,458 & 229,853\\
U2R    & 52      & 228\\
R2L    & 1,124   & 16,189\\
Total  & 494,021 & 311,029\\
\end{tabular}
\center
\label{tab:numberOfRecords}
\end{table}

\begin{table*}
\center
\caption{Results for Genetic Algorithm Experiment}
\vspace{0.20cm}
\begin{tabular}{cccccccc}
& & \multicolumn{5}{ c }{Predicted label} \\ 
& & Normal & Probe & DoS & U2R & R2L & \% Correct \\ \cline{3-8}
\multicolumn{1}{ c }{\multirow{2}{*}{Actual class} } &
\multicolumn{1}{ c| }{Normal} & 42138 & 1421 & 15835 & 486 & 713 & 69.5\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{Probe} & 398 & 2963 & 654 & 2 & 149 & 71.1   \\
\multicolumn{1}{ c  }{\multirow{2}{*}{} } &
\multicolumn{1}{ c| }{Dos} & 921 & 432 & 228489 & 1 & 10 & 99.4\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{U2R} & 146 & 21 & 8 & 43 & 10 & 18.9\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{R2L} & 11191 & 578 & 3398 & 141 & 881 & 5.4\\

@@

\begin{table*}
\center
\caption{Results for Genetic Algorithm Experiment}
\vspace{0.20cm}
\begin{tabular}{cccc}
%\cline{3-8}
& & \multicolumn{2}{ c }{Predicted label} \\ 
& & Normal & Intrusion\\ \cline{3-4}
\multicolumn{1}{ c }{\multirow{2}{*}{Actual class} } &
\multicolumn{1}{ c| }{Normal} & True Negative (42,138) & False Positive (18,455)\\
\multicolumn{1}{ c  }{}                        &
\multicolumn{1}{ c| }{Probe} & False Negative (12,528) & True Positive (237,908)\\
\end{tabular}
\center
\label{tab:genAlgResults2}
\end{table*}

@@

\begin{figure}
\centering
\caption{Fuzzy encoding for a feature}
\vspace{0.20cm}
\begin{tabular}{|cccc|} \hline
010 & 011 & 100 & 101\\
a=2 & b=3 & c=4 & d=5\\
\hline\end{tabular}
\label{fig:fuzEncodingForFeature}
\end{figure}

\begin{figure*}
\centering
\caption{A rule with 12 blocks of features}
\vspace{0.20cm}
\begin{tabular}{|cccc|c|cccc|c|} \hline
010 & 011 & 100 & 101   & ...... & 010 & 011 & 101 & 111   & DoS\\
a=2 & b=3 & c=4 & d=5   & ...... & a=2 & b=3 & c=5 & d=7   &\\ 
    &     & Block 1&    &        &     & Block 12& &       & Type\\
\hline\end{tabular}
\label{fig:rule}
\end{figure*}

@@

\begin{table*}
\center
\caption{Results from Experiment 1}
\vspace{0.20cm}
\begin{tabular}{ccccccc}
 & Attack & Normal & Total Records & FP (\%) & FN (\%) & DR (\%)\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 16000 & 26500 & 1.13 & 4.10 & 97.92\\
\end{tabular}
\label{tab:fuzGenExp1}
\center
\end{table*}

In the second experiment, seven tests were run. For each test case there were 13 attack types plus normal activity that were in the training data set. Three attack types were used for the unknown testing data set. For example, test case 1 used the training data set that does not have Advance Port Scan, Ack Scan, and Xmas Tree, which are all probe attacks. These three attacks were then used for the testing data set. Table~\ref{tab:fuzGenExp2} shows some of the results from the fuzzy genetic algorithm and a decision tree algorithm, which is another common way of addressing these problems. For further information on decision trees, see~\cite{decisionTree}. When compared with the decision tree algorithm, the fuzzy genetic algorithm has a higher detection rate in all cases except 3 and 5. It can be seen that in cases 2 and 4 the decision tree has low detection rates, while the fuzzy genetic algorithm has much higher detection rates.

\begin{table}
\caption{Unknown Attack Experiment}
\vspace{0.20cm}
\begin{tabular}{llll}
Test & Unknown & Decision & Fuzzy\\
Case & Attacks & Tree DR (\%)  & Genetic DR (\%)\\ \hline

@@

\begin{table}
\caption{KDD99 and RLD09 Results}
\vspace{0.20cm}
\begin{tabular}{cccccc}
Data set & Attack & Normal & FP (\%) & FN (\%) & DR (\%)\\ \hline
KDD99 & 160,117 & 39,337 & 0.13 & 1.55 & 98.72\\
RLD09 & 10,500 & 16,000 & 1.14 & 3.39 & 97.97\\
\end{tabular}
\label{tab:bothSetsResults}
\end{table}

@@

\begin{table}
\caption{Results for KDD99 with Certain Attacks}
\vspace{0.20cm}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & Back & DoS & 85.33 & 0.00 & 16.56\\
2 & Smurf & DoS & 0.76 & 0.10 & 99.73\\
3 & Portsweep & Probe & 6.40 & 0.00 & 93.66\\
4 & Satan & Probe & 0.74 & 3.75 & 99.22\\
\end{tabular}
\label{tab:kddAttacks}
\end{table}

@@

\begin{table}
\caption{Results for RLD99 with Certain Attacks}
\vspace{0.20cm}
\begin{tabular}{cccccc}
Test & Attack & Type & FP (\%) & FN (\%) & DR (\%)\\ \hline
1 & Smurf & DoS & 0.02 & 0 & 99.98\\
2 & UDP Flood & DoS & 11.06 & 0 & 89.59\\
3 & Ackscan & Probe & 0.03 & 0 & 99.97\\
4 & Synscan & Probe & 0.65 & 4.2 & 99.24\\
\end{tabular}
\label{tab:rldAttacks}
\end{table}
